% general_conclusion.tex
\chapter*{General Conclusion}
\addcontentsline{toc}{chapter}{General Conclusion}
\label{ch:conclusion}

\section*{Summary of Contributions}

This thesis proposed and evaluated three architectural enhancements to a reinforcement-learning-based auto-scheduler for the \gls{mlir} compiler framework. The starting point was the prior \gls{mlir}-\gls{rl} agent by Tirichine et al.~\cite{tirichine2025reinforcement}, which used an \gls{lstm}-based policy and a sparse terminal speedup reward. We identified three limitations in this baseline---hardware-agnostic observations, sequential \gls{lstm} encoding, and an exploration of whether intermediate shaped rewards could accelerate training---and addressed each.

\textbf{Hardware-aware observation.} The prior agent had no knowledge of the hardware it targeted. A policy trained on one \gls{cpu} produced suboptimal schedules when deployed on another. We introduced a 7-element hardware feature vector---L1, L2, and L3 cache sizes, physical and logical core counts, \gls{simd} width, and clock frequency---concatenated to the policy and value network inputs. This allowed the agent to condition tile sizes and parallelization decisions on the execution platform. Multi-hardware evaluation across three clusters with different \gls{cpu} architectures showed that hardware features provide their largest benefit on the training platform and degrade gracefully on dissimilar hardware, never hurting performance.

\textbf{Shaped reward exploration.} We investigated whether dense intermediate reward signals based on static code analysis---arithmetic intensity, vectorizability, and parallel-loop ratio---could accelerate training by providing feedback at each transformation step. The experimental results contradicted the initial hypothesis: the agent learned to maximize the static heuristics rather than actual execution speedup. Removing shaped reward entirely and relying solely on the terminal speedup signal produced the best-performing agent, which achieved higher speedups across the majority of model families and successfully optimized 17 out of 18 full neural network architectures. This negative result is a contribution in itself: it establishes that static heuristics are unreliable proxies for execution speedup and that the terminal signal alone is sufficient for learning effective optimization policies.

\textbf{Transformer encoder.} We replaced the sequential \gls{lstm} encoder with a Transformer that processes all loop levels simultaneously via self-attention. This allows the model to attend directly to relationships between non-adjacent loops, such as the outermost tiling loop and the innermost vectorization target. The Transformer variant provides a consistent advantage over the \gls{lstm} baseline, with the largest gains on deep loop nests where long-range dependencies between loop levels determine optimization quality.

\section*{Main Findings}

The experimental evaluation produced four findings. First, \textbf{terminal speedup alone is sufficient for learning.} The agent without reward shaping consistently outperformed all shaped-reward variants across model families. Intermediate heuristics mislead the policy because they reward properties that are necessary but not sufficient for performance---a loop nest can have high arithmetic intensity on paper while compiling to inefficient code.

Second, \textbf{hardware-aware features enable cross-platform deployment.} The hardware vector provided clear benefits on the training cluster (Bergamo, AMD EPYC 9754). On Jubail (AMD EPYC 7742), which shares the Zen microarchitecture, the hardware-aware and hardware-agnostic variants performed similarly, indicating that learned schedules transfer across architecturally similar processors. On Dalma (Intel Xeon E5-2680 v4), the hardware-aware variant maintained a modest advantage, showing that hardware features help even when the target architecture differs fundamentally from the training platform.

Third, \textbf{full neural network model optimization is feasible.} The agent successfully optimized 17 out of 18 architectures, with per-model speedups reaching 10.58$\times$ on T5, 9.62$\times$ on Llama~3.2~1B, and 8.00$\times$ on Albert. The prior agent failed to produce valid schedules on eight of these architectures. This demonstrates that learned scheduling scales from isolated operations to complete model inference pipelines.

Fourth, \textbf{the Transformer encoder provides modest but consistent gains over \gls{lstm}.} The benefit is concentrated on deep loop nests, where self-attention preserves information across non-adjacent levels. For shallow nests, the two encoders perform comparably.

\section*{Limitations}

Five limitations define the scope of this work. First, \textbf{training requires \gls{jit}-compiling every candidate transformation sequence.} A single 10{,}000-iteration training run takes approximately 70 hours on a 28-core node, making rapid experimentation impractical. Second, \textbf{static heuristics proved to be unreliable reward proxies.} The metrics used for reward shaping do not correlate well with execution speedup across diverse operation types, motivating their removal. A learned cost model that predicts execution time from static features could replace both the heuristics and the expensive compilation step. Third, \textbf{evaluation is limited to \gls{cpu} targets.} \gls{gpu} and accelerator backends introduce distinct optimization priorities---memory coalescing, shared memory usage, warp occupancy---that require different transformation strategies and hardware features. Fourth, the \textbf{action space} is fixed to five transformation types; loop unrolling, padding, and packing remain unexplored. Fifth, the policy makes \textbf{greedy decisions at inference time} without the ability to backtrack or explore alternative sequences.

\section*{Future Work}

Several directions follow from these limitations. A \textbf{learned cost model} that predicts execution speedup from the agent's observation could replace both the static heuristics and the expensive \gls{jit} compilation step for intermediate rewards, decoupling policy optimization from the compilation bottleneck. Extending the system to \textbf{\gls{gpu} and accelerator backends} through \gls{mlir}'s GPU dialect would test whether the same \gls{rl} framework generalizes to a different optimization domain. \textbf{Expanding the action space} with loop unrolling, padding, and packing would unlock additional speedup opportunities for specific operation patterns. \textbf{Search-based inference} with beam search or Monte Carlo Tree Search~\cite{tirichine2025reinforcement} could improve final schedule quality by evaluating multiple candidate sequences, at the cost of additional inference-time computation.

\section*{Closing Remarks}

This thesis systematically evaluated three independent enhancements to \gls{rl}-based compiler optimization---hardware-aware observations, shaped reward exploration, and attention-based state encoding---and provided empirical evidence for the contribution of each. The finding that terminal speedup alone outperforms intermediate shaped rewards is a practical lesson for building \gls{rl}-based optimizers: proxy metrics must be validated against the true objective before being used as training signals. The finding that hardware features enable zero-shot transfer across \gls{cpu} architectures points toward compilers that adapt their optimization strategies to the deployment target without manual tuning.

This work extends a research trajectory that began with Bendib et al.~\cite{bendib2024reinforcement} and continued through Tirichine et al.~\cite{tirichine2025reinforcement}. Each iteration has expanded the scope of learned scheduling---from single-operation optimization to transformer-based encoding to cross-platform deployment. The remaining challenges---replacing heuristic rewards with learned cost models, extending to accelerator backends, and integrating search-based inference---represent natural next steps toward a compiler that learns to optimize any program for any hardware.
