\chapter*{General Introduction}
\addcontentsline{toc}{chapter}{General Introduction}
\label{ch:introduction}

Code optimization is a fundamental area of computer science concerned with transforming programs to improve their performance, whether measured in execution time, memory usage, energy consumption, or a combination of these metrics. At the compiler level, optimization translates high-level algorithmic intent into machine instructions that exploit the full capabilities of the underlying hardware. As software workloads grow in scale and complexity, and as hardware architectures diversify from multi-core CPUs to GPUs and specialized accelerators, the task of producing efficient code has become increasingly challenging. Traditionally, compilers rely on hand-crafted heuristics designed by domain experts: rules that dictate when to apply loop tiling, which tile sizes to use, whether to fuse two loops, or in which order to apply transformations. These heuristics require extensive manual effort to develop and tune for each new architecture. Furthermore, they are inherently fragile: a rule that works well for one class of programs may perform poorly on another, and a schedule optimized for a server-class processor rarely transfers to an embedded device. The \textbf{phase ordering problem}, which refers to the fact that transformations interact in non-linear ways and that the optimal sequence depends on both the program and the target hardware, has persisted as one of the central open challenges in compiler design~\cite{touati2014phase}. These limitations have driven a growing interest in automated, data-driven approaches to compiler optimization.

Reinforcement learning (RL) is a machine learning paradigm in which an agent learns to make sequential decisions by interacting with an environment and receiving feedback in the form of a numerical reward~\cite{sutton2018reinforcement}. Unlike supervised learning, which requires a dataset of correct answers, RL discovers effective strategies through trial and error, guided only by the reward signal. This makes RL particularly well suited to compiler optimization, where the ``correct'' sequence of transformations is not known in advance and can only be assessed empirically by compiling and executing the resulting code. Deep reinforcement learning extends this framework by using deep neural networks as function approximators, enabling RL agents to operate in high-dimensional and complex state spaces. Over the past decade, deep RL has produced breakthrough results in domains such as game playing~\cite{mnih2015human}, robotics~\cite{levine2016end}, and chip design~\cite{mirhoseini2021placement}, demonstrating the ability to learn sophisticated strategies that surpass human-designed heuristics. In the context of compilers, RL offers a path toward adaptive, architecture-aware scheduling that can discover optimization sequences not anticipated by compiler engineers.

The Multi-Level Intermediate Representation (MLIR) is a compiler infrastructure within the LLVM ecosystem, introduced by Lattner et al.~\cite{lattner2021mlir}, that addresses a key limitation of traditional compiler intermediate representations: the assumption of a single, fixed level of abstraction. MLIR instead supports a \textit{multi-level} IR where different \textit{dialects} define operations, types, and attributes at different levels of abstraction, all coexisting within the same representation. This design allows high-level domain-specific operations to be progressively lowered to low-level machine instructions through a series of well-defined passes. Of particular relevance to this work is the Linalg dialect, which represents structured linear algebra operations, and the Transform dialect, which encodes transformation schedules as operations that can be applied, composed, and replayed programmatically. These properties make MLIR an attractive platform for RL-based optimization: the explicit loop structure captured by Linalg operations provides a rich state representation, the Transform dialect serves as a natural action interface, and MLIR's Python bindings enable the entire RL training pipeline to be implemented in a unified Python environment. The broader vision driving this line of research is to develop RL-guided schedulers that generalize across programs and hardware targets within a unified compiler framework, reducing the reliance on hand-tuned heuristics and making high-performance compilation accessible to a wider range of applications.

The application of RL to MLIR-based optimization was initiated by Bendib et al.~\cite{bendib2024reinforcement}, who developed a reinforcement learning environment for optimizing individual Linalg operations. Their agent selected and parameterized loop transformations (tiling, parallelization, interchange, vectorization) using an LSTM-based policy network, and was rewarded based on the execution-time speedup achieved. Building on this foundation, Tirichine et al.~\cite{tirichine2025reinforcement} extended the environment to support full neural network computation graphs. Their system introduced an abstract syntax tree (AST) based state representation with a recursive LSTM encoder to capture hierarchical producer-consumer relationships, a Fusion action that merges operation pairs, and a dataset of synthesized sequences and blocks extracted from convolutional neural network models. Their evaluation on VGG, ResNet, and MobileNet demonstrated that the RL agent could achieve notable speedups, including a 2.5$\times$ improvement on MobileNetV2 relative to standard PyTorch execution. However, their analysis also identified several limitations that constrain the system's optimization capability. First, the agent's observation was hardware-agnostic: it received no information about the target processor's cache hierarchy, core count, or SIMD width, forcing it to infer platform-appropriate strategies through costly trial and error. Second, the reward function was entirely sparse, providing feedback only at the end of an episode and offering no guidance on intermediate transformation choices. Third, the recursive LSTM encoder processes loop levels sequentially, which can limit its ability to capture long-range dependencies between non-adjacent levels in deeply nested loops. Fourth, the action space was restricted to six transformation types, excluding operations such as padding, packing, and loop unrolling that can improve memory alignment and instruction-level parallelism. These limitations define the research space that the present work addresses.

This thesis develops a set of complementary improvements to the RL-based MLIR auto-scheduler, organized into four incremental versions that culminate in a unified system (V4). Version V1 introduces a \textbf{hardware-aware observation space} that encodes the target processor's cache sizes, core count, SIMD width, and clock frequency directly into the agent's input, enabling platform-specific optimization decisions. Version V2 replaces the sparse terminal reward with a \textbf{shaped reward function} that supplements the execution-time speedup with dense intermediate rewards computed from static code analysis, providing learning signal at every transformation step. Version V3 replaces the recursive LSTM encoder with a \textbf{transformer-based loop nest encoder} that uses structured tokenization and multi-head self-attention, allowing the model to attend directly to relationships between any pair of loop levels regardless of their distance in the nest hierarchy. Version V4 combines all three improvements and further introduces an \textbf{expanded action space} with pad, pack, and unroll transformations, giving the agent finer control over data layout and instruction-level parallelism. Each version is developed and evaluated independently to measure the contribution of individual design choices, and a comparative dashboard is provided to analyze performance across versions. The remainder of this manuscript is organized as follows. Chapter~1 provides the background on compiler code optimization, covering the compilation pipeline, loop nest transformations, and the MLIR infrastructure. Chapter~2 introduces the foundations of reinforcement learning, from Markov decision processes through policy gradient methods and Proximal Policy Optimization. Chapter~3 surveys related work on applying RL to compiler optimization, concluding with a comparative synthesis. Chapter~4 presents the design and implementation of our system, detailing the observation space, action space, reward function, and policy architecture for each version. Chapter~5 describes the experimental methodology, results, and analysis. A general conclusion summarizes our findings and discusses directions for future work.
